### 2022-08-11 draft.tex
Deploying to gh-pages from @ FrancisBanville/ms_maxent_networks@65f93fed31c2742c2277d044da061b61287dd750 🚀

Food-web properties that can be derived using MaxEnt are varied and
pertain to different elements of the network (i.e.~at the species
(node), the interaction (edge) or the community (network) levels).
Because MaxEnt is a method of finding least-biased probability
distributions given partial knowledge about a system, these properties
need to be represented probabilistically. For example, at the species
level, MaxEnt can be used to predict the distribution of trophic levels
among species, as well as the distribution of species' vulnerability
(number of predators) and generality (number of preys). By contrast, at
the interaction level, predictions can be made on the distribution of
interaction strengths in weighted food webs. At the community level, it
can generate probability distributions of many measures of their
emerging structure and of networks themselves (i.e.~a probability
distribution that specific network configurations are realized given the
model and constraints). Overall, the potential of this method in the
study of food webs is broad. The applicability and performance of MaxEnt
mostly depend on the ecological information available and on our
capacity to find the right set of state variables that best represent
natural systems and to translate them into appropriate statistical
constraints. Having a validated maximum entropy model for the system at
hand allows us to make least-biased predictions using a minimal amount
of data, as well as identify the most important ecological processes
shaping that system. In other words, MaxEnt is a valuable and
underexploited tool in network ecology that allows us to better
understand and predict the structure of ecological networks.

@@

The joint degree distribution \(p(k_{in},k_{out})\) of a food web with
\(S\) species is a joint discrete probability distribution describing
the probability that a species has \(k_{in}\) predators and \(k_{out}\)
preys, with \(k_{in}\) and \(k_{out}\) \(\epsilon\) \([0, S]\). Basal
species (e.g., plants) have a \(k_{out}\) of \(0\), whereas top
predators have a \(k_{in}\) of \(0\). In contrast, the maximum number of
preys and predators a species can have is set by the number of species
in the food web. Here we show how the joint degree distribution of
maximum entropy can be obtained given knowledge of the number of species
\(S\) and the number of interactions \(L\).

@@

One aspect of the degree distribution that informs us of its ecological
realism is the number of isolated species it predicts. As MacDonald,
Banville, and Poisot (2020) pointed out, the size of food webs should at
least be of \(S-1\) interactions, since a lower number would yield
isolated species, i.e.~species without any predators or preys. Because
non-basal species must eat to survive, isolated species could indicate
that other species are missing or they could simply be removed from the
food web. In fig.~S2, we show that the degree distribution of maximum
entropy, given \(S\) and \(L\), gives very low probabilities that a
species will be isolated in its food web (i.e.~having \(k = 0\)) above
the \(S-1\) threshold. However, under our purely information-theoretic
model, the probability that a species is isolated is quite high when the
total number of interactions is below \(S-1\). Moreover, the expected
proportion of isolated species rapidly declines by orders of magnitude
with increasing numbers of species and interactions (fig.~S2). This
supports the ecological realism of the degree distribution of maximum
entropy derived above. Nevertheless, ecologists wanting to model a
system without allowing isolated species could simply change the lower
limit of \(k\) to \(1\) in eq.~\ref{eq:lagrange2_dd} and solve the
resulting equation numerically.

Using the analytical approach, we showed how important measures of
food-web structure, namely the degree distribution and the joint degree
distribution, could be derived with the principle of maximum entropy
given minimal knowledge about a biological community. This type of
models, although useful to make least-biased predictions on many network
properties, can be hard to apply for other measures. Indeed, there are
dozens of measures of network structure (Delmas et al. 2019) and many
are not calculated with mathematical equations, but with algorithms.
Moreover, the applicability of this method to empirical systems is
limited by the state variables we can actually measure and use. In the
next section, we propose a more flexible method to predict many measures
of network structure simultaneously, i.e.~by finding heuristically the
network configuration having maximum entropy given partial knowledge of
its emerging structure.

@@

We generated probabilistic networks using both types of null models for
all empirical food webs in our complete dataset. Then, we converted
these networks to adjacency matrices of Boolean values by generating
\(100\) random networks for each of these probabilistic webs. Finally,
we kept the \(L\) entries that were sampled the most amount of times,
with \(L\) given by the number of interactions in each food web. This
ensured that the resulting null networks had the same number of
interactions as their empirical counterparts. Thus, for each null model,
we ended up with one null adjacency matrix for each empirical network.

@@

where \(n_i\) and \(n_j\) are the abundances (or biomass) of both
species, and \(N\) is the total abundance (or biomass) of all species in
the network. We predicted neutral abundance matrices for all empirical
food webs in our abundance dataset, and converted these weighted
networks to adjacency matrices of Boolean values using the same method
as the one we used for our null models.

@@

@@

The principle of maximum entropy can be applied on the network itself if
we decompose its adjacency matrix into a non-zero vector of relative
values. This is a necessary step when working with food webs, which are
frequently expressed as a matrix \(A = [a_{ij}]\) of Boolean values
representing the presence (\(a_{ij} = 1\)) or absence (\(a_{ij} = 0\))
of an interaction between two species \(i\) and \(j\). Knowing one or
many properties of a food web of interest (e.g., its number of species
and number of interactions), we can simulate its adjacency matrix
randomly by using these known ecological information to constrain the
space of potential networks. The entropy of this hypothetical matrix can
then be measured after decomposing it into appropriate values.
Simulating a series of networks until we find the one having the highest
entropy allows us to search for the most complex food-web configuration
given the ecological constraints used. This configuration is the least
biased one considering the information we had. In other words, the most
we can say about a network's adjacency matrix, without making more
assumptions than the ones given by our incomplete knowledge of its
structure, is the one of maximum entropy. Generating the most complex
network that corresponds to this structure allows us to explore more
easily other properties of food webs under MaxEnt.

Shannon's entropy can only be calculated on conventional probability
distributions such as the joint degree distribution. This is an issue
when working with ecological networks, which are represented as
adjacency matrices. For this reason, we need to use another measure of
entropy if we want to predict the network itself. We used the SVD
entropy as our measure of entropy, which is an application of Shannon's
entropy to the relative non-zero singular values of a truncated singular
value decomposition (t-SVD; Strydom, Dalla Riva, and Poisot 2021) of a
food web's Boolean adjacency matrix. We measured SVD entropy as follows:

@@

where \(s_i\) are the relative singular values of the adjacency matrix
(\(s_i = \sigma_i / \sum_{i = 1}^R \sigma_i\), where \(\sigma_i\) are
the singular values). Note that the distribution of relative singular
values is analogous to a probability distribution, with \(0 < s_i < 1\)
and \(\sum s_i = 1\). Thus, this measure also satisfies all four
properties of an appropriate entropy measure (see introduction), while
being a proper measure of the internal complexity of food webs (Strydom,
Dalla Riva, and Poisot 2021). Following Strydom, Dalla Riva, and Poisot
(2021), we standardized this measure with the rank \(R\) of the matrix
(i.e.~\(J / \ln(R)\)) to account for the difference in dimensions
between networks (\emph{sensu} Pielou's evenness; Pielou 1975). We used
SVD entropy to predict the network configuration of maximum entropy
(i.e.~of maximum complexity) heuristically given a predetermined set of
constraints.

\hypertarget{maxent-network-models-types-i-and-ii}{%
\subsection{MaxEnt network models (types I and
II)}\label{maxent-network-models-types-i-and-ii}}

We built two types of MaxEnt network models: one based on connectance
(type I MaxEnt network model) and the other based on the joint degree
sequence (type II MaxEnt network model). They are based on the same
constraints as the types I and II null models presented above. For both
models, we used a simulated annealing algorithm with \(4\) chains,
\(2000\) steps and an initial temperature of \(0.2\). For each chain, we
first generated one random Boolean matrix with the same order (number of
species) as empirical webs, while maintaining the total number of
interactions (type I MaxEnt network model) or rows and columns sums
(type II MaxEnt network model). These are our initial configurations.
Then, we swapped interactions sequentially while maintaining the
original connectance or the joint degree sequence for types I and II
MaxEnt network model, respectively. Configurations with a higher SVD
entropy than the previous one in the chain were always accepted, whereas
they were accepted with a probability conditional to a decreasing
temperature when lower. The final configuration with the highest SVD
entropy among the four chains constitute our estimated MaxEnt network.
Even though we decided to work with point estimates, it is possible to
have a (non MaxEnt) probability distribution of networks when working
with the entire chains (or part of the chains). For each network in our
complete and abundance datasets, we estimated their configuration with
maximum entropy using both types of MaxEnt network models.

@@

We compared different measures of the structure of empirical food webs
with the ones of null, neutral, and MaxEnt networks
tbl.~\ref{tbl:measures_all}, and \ref{tbl:measures_abund}. For instance,
we measured the nestedness \(\rho\) according to Staniczenko, Kopp, and
Allesina (2013), i.e.~using the spectral radius of the adjacency matrix.
Nestedness indicates how much the diet of specialist species is a subset
of the one of generalists, and is strongly associated with network
modularity (Fortuna et al. 2010). We also measured the maximum trophic
level \emph{maxtl}, network diameter \emph{diam} (i.e.~the longest of
the shortest paths between all species pairs; Albert and Barabasi 2002),
the average maximum similarity between species pairs \emph{MxSim}, the
proportion of cannibal species \emph{Cannib} (i.e.~the proportion of
self loops), and the proportion of omnivorous species \emph{Omniv}
(i.e.~species whose preys are of different trophic levels).
\emph{MxSim}, \emph{Cannib}, and \emph{Omniv} are more deeply defined in
Williams and Martinez (2000).

@@

The principle of maximum entropy is a robust mathematical method of
finding least-biased probability distributions that have some desired
properties given by prior knowledge on a system. We first used MaxEnt to
predict the joint degree distribution and the degree distribution of
maximum entropy given known numbers of species and of interactions. We
found that the resulting joint degree distributions were more symmetric
than the ones of empirical food webs, which suggests that other
constraints might be needed to improve those predictions. We also used
MaxEnt to predict networks of maximum entropy with a specified
structure. These networks are the most complex, or random, given the
constraints used. Likewise, we found that knowledge of species richness
and of the number of interactions were not sufficient to reproduce many
aspects of network structure found in nature. However, a model based on
the entire joint degree sequence, i.e.~on the number of preys and
predators for each species, gave more convincing results. Our type II
MaxEnt network model yielded better or similar predictions than the type
II null model, also based on the joint degree sequence, for most
measures considered including the motifs profile. This suggests that the
joint degree sequence drives many aspects of network structure. Indeed,
considering our findings that empirical networks are close to maximum
entropy for a given joint degree sequence, our results suggest that
food-web topology is configured almost entirely randomly around these
marginal numbers of predators and preys.

@@

@@

One of the biggest challenges in using the principle of maximum entropy
is to identify the set of state variables that best reproduce empirical
data. We found that the numbers of preys and predators for each species
are important state variables for the prediction of maximum entropy
networks. However, our predictions overestimated some measures of
network structure, especially the maximum trophic level and network
diameter. Therefore, we should continue playing the ecological detective
to find these other topological constraints that would improve the
predictions of our MaxEnt network models.

>>>>

Food-web properties that can be derived using MaxEnt are varied and
pertain to different elements of the network (i.e.~at the species
(node), the interaction (edge) or the community (network) levels).
Because MaxEnt is a method of finding least-biased probability
distributions given partial knowledge about a system, these properties
need to be represented probabilistically. For example, at the species
level, MaxEnt can be used to predict the distribution of trophic levels
among species, as well as the distribution of species' vulnerability
(number of predators) and generality (number of prey). By contrast, at
the interaction level, predictions can be made on the distribution of
interaction strengths in weighted food webs. At the community level, it
can generate probability distributions of many measures of their
emerging structure and of networks themselves (i.e.~a probability
distribution that specific network configurations are realized given the
model and constraints). Overall, the potential of this method in the
study of food webs is broad. The applicability and performance of MaxEnt
mostly depend on the ecological information available and on our
capacity to find the right set of state variables that best represent
natural systems and to translate them into appropriate statistical
constraints. Having a validated maximum entropy model for the system at
hand allows us to make least-biased predictions using a minimal amount
of data, as well as identify the most important ecological processes
shaping that system. In other words, MaxEnt is a valuable and
underexploited tool in network ecology that allows us to better
understand and predict the structure of ecological networks.

@@

The joint degree distribution \(p(k_{in},k_{out})\) of a food web with
\(S\) species is a joint discrete probability distribution describing
the probability that a species has \(k_{in}\) predators and \(k_{out}\)
prey, with \(k_{in}\) and \(k_{out}\) \(\epsilon\) \([0, S]\). Basal
species (e.g., plants) have a \(k_{out}\) of \(0\), whereas top
predators have a \(k_{in}\) of \(0\). In contrast, the maximum number of
prey and predators a species can have is set by the number of species in
the food web. Here we show how the joint degree distribution of maximum
entropy can be obtained given knowledge of the number of species \(S\)
and the number of interactions \(L\).

@@

One aspect of the degree distribution that informs us of its ecological
realism is the number of isolated species it predicts. As MacDonald,
Banville, and Poisot (2020) pointed out, the size of food webs should at
least be of \(S-1\) interactions, since a lower number would yield
isolated species, i.e.~species without any predators or prey. Because
non-basal species must eat to survive, isolated species could indicate
that other species are missing or they could simply be removed from the
food web. In fig.~S2, we show that the degree distribution of maximum
entropy, given \(S\) and \(L\), gives very low probabilities that a
species will be isolated in its food web (i.e.~having \(k = 0\)) above
the \(S-1\) threshold. However, under our purely information-theoretic
model, the probability that a species is isolated is quite high when the
total number of interactions is below \(S-1\). Moreover, the expected
proportion of isolated species rapidly declines by orders of magnitude
with increasing numbers of species and interactions. This supports the
ecological realism of the degree distribution of maximum entropy derived
above. Nevertheless, ecologists wanting to model a system without
allowing isolated species could simply change the lower limit of \(k\)
to \(1\) in eq.~\ref{eq:lagrange2_dd} and solve the resulting equation
numerically.

@@

We generated probabilistic networks using both types of null models for
all empirical food webs in our complete dataset. Then, we converted
these networks to adjacency matrices of Boolean values by generating
\(100\) random networks for each of these probabilistic webs, and kept
the \(L\) entries that were sampled the most amount of times, with \(L\)
given by the number of interactions in each food web. This ensured that
the resulting null networks had the same number of interactions as their
empirical counterparts. Thus, for each null model, we ended up with one
null adjacency matrix for each empirical network.

@@

where \(n_i\) and \(n_j\) are the abundances (or biomass) of both
species, and \(N\) is the total abundance (or biomass) of all species in
the network. We generated neutral abundance matrices for all empirical
food webs in our abundance dataset, and converted these weighted
networks to adjacency matrices of Boolean values using the same method
as the one we used for our null models.

@@

With the analytical approach, we showed how important measures of
food-web structure (e.g., the degree distribution and the joint degree
distribution) can be derived with the principle of maximum entropy using
minimal knowledge about a biological community. This type of models,
although useful to make least-biased predictions of many network
properties, can be hard to apply for other measures. Indeed, there are
dozens of measures of network structure (Delmas et al. 2019) and many
are not directly calculated with mathematical equations, but are instead
estimated algorithmically. Moreover, the applicability of this method to
empirical systems is limited by the state variables we can actually
measure and use. Here, we propose a more flexible method to predict many
measures of network structure simultaneously, i.e.~by finding
heuristically the network configuration having maximum entropy given
partial knowledge of its emerging structure.

@@

The principle of maximum entropy can be applied on the network itself if
we decompose its adjacency matrix into a non-zero vector of relative
values. This is a necessary step when working with food webs, which are
frequently expressed as a matrix \(A = [a_{ij}]\) of Boolean values
representing the presence (\(a_{ij} = 1\)) or absence (\(a_{ij} = 0\))
of an interaction between two species \(i\) and \(j\). Knowing one or
many properties of a food web of interest (e.g., its number of species
and number of interactions), we can simulate its adjacency matrix
randomly by using these known ecological information to constrain the
space of potential networks. The entropy of this hypothetical matrix can
then be measured after decomposing it into appropriate values.
Simulating a series of networks until we find the one having the highest
entropy allows us to search for the most complex food-web configuration
given the ecological constraints used. This configuration is the least
biased one considering the information available. In other words, the
most we can say about a network's adjacency matrix, without making more
assumptions than the ones given by our incomplete knowledge of its
structure, is the one of maximum entropy. Generating the most complex
network that corresponds to this structure allows us to explore more
easily other properties of food webs under MaxEnt.

Shannon's entropy can only be calculated on conventional probability
distributions such as the joint degree distribution. This is an issue
when working with the adjacency matrix of ecological networks. For this
reason, we need to use another measure of entropy if we want to predict
a network's configuration directly using MaxEnt. We used the SVD entropy
as our measure of entropy, which is an application of Shannon's entropy
to the relative non-zero singular values of a truncated singular value
decomposition (t-SVD; Strydom, Dalla Riva, and Poisot 2021) of a food
web's Boolean adjacency matrix. We measured SVD entropy as follows:

@@

where \(s_i\) are the relative singular values of the adjacency matrix
(\(s_i = \sigma_i / \sum_{i = 1}^R \sigma_i\), where \(\sigma_i\) are
the singular values). Note that the distribution of relative singular
values is analogous to a probability distribution, with \(0 < s_i < 1\)
and \(\sum s_i = 1\). This measure also satisfies all four properties of
an appropriate entropy measure above-mentioned, while being a proper
measure of the internal complexity of food webs (Strydom, Dalla Riva,
and Poisot 2021). Following Strydom, Dalla Riva, and Poisot (2021), we
standardized this measure with the rank \(R\) of the matrix
(i.e.~\(J / \ln(R)\)) to account for the difference in dimensions
between networks (\emph{sensu} Pielou's evenness; Pielou 1975).

\hypertarget{types-i-and-ii-heuristic-maxent-models}{%
\subsection{Types I and II heuristic MaxEnt
models}\label{types-i-and-ii-heuristic-maxent-models}}

We used SVD entropy to predict the network configuration of maximum
entropy (i.e.~of maximum complexity) heuristically given different
constraints for all food webs in our complete dataset. We built two
types of heuristic MaxEnt models that differ on the constraint used. The
type I heuristic MaxEnt model is based on connectance, whereas the type
II heuristic MaxEnt model is based on the joint degree sequence. These
models are thus based on the same constraints as the types I (Fortuna
and Bascompte 2006) and II (Bascompte et al. 2003) null models (Box 2)
frequently used to generate random networks topologically. This allows
direct comparison of the performance of null and heuristic MaxEnt models
in reproducing the emerging structure of empirical food webs.

For each network in our complete dataset, we estimated their
configuration of maximum entropy given each of these constraints. For
both types of heuristic MaxEnt models, we used a simulated annealing
algorithm with \(4\) chains, \(2000\) steps and an initial temperature
of \(0.2\). For each food web, we first generated one random Boolean
matrix per chain while fixing the number of species. We also maintained
the total number of interactions (i.e.~the sum of all elements in the
matrix) in the type I MaxEnt model and the joint degree sequence
(i.e.~the rows and columns sums) in the type II MaxEnt model. These were
our initial configurations. Then, we swapped interactions sequentially
while maintaining the original connectance or joint degree sequence.
Configurations with a higher SVD entropy than the previous one in the
chain were always accepted, whereas they were accepted with a
probability conditional to a decreasing temperature when lower. The
final configuration with the highest SVD entropy among the four chains
constitute the estimated maximum entropy configuration of a food web
given the constraint used.

@@

We compared different measures of the structure of empirical food webs
with the ones of null, neutral, and MaxEnt networks
tbl.~\ref{tbl:measures_all}, and \ref{tbl:measures_abund}. For instance,
we measured the nestedness \(\rho\) according to Staniczenko, Kopp, and
Allesina (2013), i.e.~using the spectral radius of the adjacency matrix.
Nestedness indicates how much the diet of specialist species is a subset
of the one of generalists, and is strongly associated with network
modularity (Fortuna et al. 2010). We also measured the maximum trophic
level \emph{maxtl}, network diameter \emph{diam} (i.e.~the longest of
the shortest paths between all species pairs; Albert and Barabasi 2002),
the average maximum similarity between species pairs \emph{MxSim}, the
proportion of cannibal species \emph{Cannib} (i.e.~the proportion of
self loops), and the proportion of omnivorous species \emph{Omniv}
(i.e.~species whose prey are of different trophic levels). \emph{MxSim},
\emph{Cannib}, and \emph{Omniv} are more deeply defined in Williams and
Martinez (2000).

@@

The principle of maximum entropy is a robust mathematical method of
finding least-biased probability distributions that have some desired
properties given by prior knowledge on a system. We first used MaxEnt to
predict the joint degree distribution and the degree distribution of
maximum entropy given known numbers of species and of interactions. We
found that the resulting joint degree distributions were more symmetric
than the ones of empirical food webs, which suggests that other
constraints might be needed to improve those predictions. We also used
MaxEnt to predict networks of maximum entropy with a specified
structure. These networks are the most complex, or random, given the
constraints used. Likewise, we found that knowledge of species richness
and of the number of interactions were not sufficient to reproduce many
aspects of network structure found in nature. However, a model based on
the entire joint degree sequence, i.e.~on the number of prey and
predators for each species, gave more convincing results. Our type II
MaxEnt network model yielded better or similar predictions than the type
II null model, also based on the joint degree sequence, for most
measures considered including the motifs profile. This suggests that the
joint degree sequence drives many aspects of network structure. Indeed,
considering our findings that empirical networks are close to maximum
entropy for a given joint degree sequence, our results suggest that
food-web topology is configured almost entirely randomly around these
marginal numbers of predators and prey.

@@

Even though we decided to work with point estimates, it is possible to
have a (non MaxEnt) probability distribution of networks when working
with the entire chains (or part of the chains).

@@

One of the biggest challenges in using the principle of maximum entropy
is to identify the set of state variables that best reproduce empirical
data. We found that the numbers of prey and predators for each species
are important state variables for the prediction of maximum entropy
networks. However, our predictions overestimated some measures of
network structure, especially the maximum trophic level and network
diameter. Therefore, we should continue playing the ecological detective
to find these other topological constraints that would improve the
predictions of our MaxEnt network models.
